Probabilistic bioregion models suggest seabird assemblages are linked to the boundary current off eastern Australia

### 2023-06-09 docs/ms-draft.Rmd
found a bug and needed to re-run results, including some new files as well

---
title: "Probabilistic bioregion models suggest seabird assemblages are linked to the main, fast-changing oceanic current off eastern Australia"
# "Probabilistic bioregion models suggest seabird assemblages are linked to large scale oceanography off eastern Australia"
# author: Nicholas W Daudt
# date: '`r format(Sys.Date())`'
author:
  - Nicholas W. Daudt:
      email: nicholaswdaudt@gmail.com
      institute: [MarSciUO]
      correspondence: true
  - Eric J. Woehler:
      institute: [BirdLife, IMAS]
      correspondence: false
  - Matthew R. Schofield:
      institute: [MathStatUO]
      correspondence: false
  - Robert O. Smith:
      institute: [MarSciUO]
      correspondence: false
  - Leandro Bugoni:
      institute: [FURG]
      correspondence: false
  - William J. Rayment:
      institute: [MarSciUO]
      correspondence: false
institute:
  - MarSciUO: Department of Marine Science, University of Otago, Aotearoa New Zealand
  - BirdLife: Australasian Seabird Group, BirdLife Australia, Hobart, Australia
  - IMAS: Institute Marine and Antarctic Studies, University of Tasmania, Hobart, Australia
  - MathStatUO: Department of Mathematics and Statistics, University of Otago, Aotearoa New Zealand
  - FURG: Institute of Biological Sciences, Universidade Federal do Rio Grande--FURG, Rio Grande, RS, Brazil
abstract: |
  **Aim**: To identify assemblages of seabird species at large scale, probabilistically, and related them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used seabird at-sea ship-based survey data collected over 37° latitude and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and counts). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within it but different from other regions. These groups also vary according to covariates, on which we included oceanographic static and climatic, as well as physiographic layers.
  **Results**: All models suggested two large scale assemblages ('northern' and 'southern' assemblages), except for the abundance-autumn model which indicated three groups. Model results seem to agree in a biogeographic barrier at ~32°S, where the East Australian Current mostly dissipates to the east and loses force. Furthermore, sea surface temperature and, in one case, sea surface salinity, were selected in all final models, further indicating close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and counts resulted in similar spatial and species profile patterns.
  **Conclusion**: Two seabird assemblages were clearly identified from RCP models, suggesting the persistence of these groups at temporal and spatial scales. Given the rapid change that the East Australian Current is experiencing, and its importance on driving seabirds' distributions, the methods applied in our study could be replicated in future analyses to assess possible spatial/temporal changes in seabird assemblages, including on its structure.
output: 
  # html_document: default
  # word_document:
  #   reference_docx: "../docs/docx-template.docx"
  # officedown::rdocx_document:
  #   reference_docx: "../docs/docx-template.docx"
  # rticles::sim_article:
  #   keep_tex: TRUE
  bookdown::word_document2:
    pandoc_args:
      - --lua-filter=../docs/scholarly-metadata.lua
      - --lua-filter=../docs/author-info-blocks.lua
      - --reference-doc=../docs/docx-template.docx 
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
---

@@

Seabird data were gathered aboard RV Investigator and RV Falkor during 15 voyages, between 2016--2021, over \~37° of latitude (**Fig. 1b**; **Table SX**). One observer continuously recorded seabirds within a 300-m forward quadrant (eye height **18-m, Eric?** a.s.l) throughout daylight hours while the ship was underway at a constant speed. Species, the number of birds and behaviour were logged directly onto a data entry system through a ship-based web server. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species) **(Eric: is there any ref for the protocol?)**.

@@

knitr::include_graphics("../EDA/Fig1-study-area-raw-counts.png")
```

@@

All models suggested two RCPs as the optimal grouping number, except for P/A-autumn, which indicated three RCPs as the best fit (**Fig S3. - multifit plots**). The best fitting models generally had **1--3** covariates selected, although the abundance-autumn model had five covariates (**Table 2 - full-model/selected model**). For the P/A-autumn model, we have deliberately chosen the 'second' best model based on parsimony---the top model had six covariates but was \<1 BIC unit lower than the second model with only two covariates. All the best models selected SST or SSS, variables essential for characterising water masses. The best models show a good fit with no apparent deviances in assumptions (**Fig. S4 - residuals**).

@@

<!-- Although we did not predict the models onto a future projection, a note on possible climate change effects on seabird off eastern Australia must be done.  -->

@@

While remote tracking studies are advancing our understanding on how seabirds use their environment at fine-, individual-scale level [@tremblay2009], they fail to perceive how groups of species (assemblages) relate to the environment and each other on a shared seascape. Therefore, using at-sea counts of seabirds is a well-established method to capture these relationships [@ainley2012; @woehler2003; @hyrenbach2007]. On the other hand, the spatial scale on which assemblages relate to their environment is also a field of debate [@hunt1987; @ribic1997; @scales2017]. Understanding the biology of the studied group is thus essential to set the appropriate scale of analysis. Our choice of 1° latitude $\times$ 1° longitude was mainly influenced by data availability and the trade-off between number of species per grid/season (see also \@ref(limitations)); nonetheless, the chosen spatial scale is similar with seabird distribution analyses elsewhere [@harris2007; @santora2013; @sojitra2022]. Therefore, for describing seabird assemblages at a large scale, we believe the chosen spatial scale was a good fit, and models described **expected?** patterns of species composition.

>>>>

---
title: "Probabilistic bioregion models suggest seabird assemblages are linked to the boundary current off eastern Australia"
# "Probabilistic bioregion models suggest seabird assemblages are linked to large scale oceanography off eastern Australia"
# author: Nicholas W Daudt
# date: '`r format(Sys.Date())`'
author:
  - Nicholas W. Daudt:
      email: nicholaswdaudt@gmail.com
      institute: [MarSciUO]
      correspondence: true
  - Eric J. Woehler:
      institute: [ASG]
      correspondence: false
  - Matthew R. Schofield:
      institute: [MathStatUO]
      correspondence: false
  - Robert O. Smith:
      institute: [MarSciUO]
      correspondence: false
  - Leandro Bugoni:
      institute: [FURG]
      correspondence: false
  - William J. Rayment:
      institute: [MarSciUO]
      correspondence: false
institute:
  - MarSciUO: Department of Marine Science, University of Otago, Aotearoa New Zealand
  - ASG: Australasian Seabird Group, BirdLife Australia, Hobart, Australia
  - MathStatUO: Department of Mathematics and Statistics, University of Otago, Aotearoa New Zealand
  - FURG: Institute of Biological Sciences, Universidade Federal do Rio Grande--FURG, Rio Grande, RS, Brazil
abstract: |
  **Aim**: To identify assemblages of seabird species at large scale, probabilistically, and related them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used seabird at-sea ship-based survey data collected over 37° latitude and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and counts). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within it but different from other regions. These groups also vary according to covariates, on which we included oceanographic static and climatic, as well as physiographic layers.
  **Results**: All models suggested two large scale assemblages ('northern' and 'southern' assemblages), except for the abundance-autumn model which indicated three groups. Model results seem to agree in a biogeographic barrier at ~32°S, where the East Australian Current mostly dissipates to the east and loses force. Furthermore, sea surface temperature and, in one case, sea surface salinity, were selected in all final models, further indicating close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and counts resulted in similar spatial and species profile patterns.
  **Conclusion**: Two seabird assemblages were clearly identified from RCP models, suggesting the persistence of these groups at temporal and spatial scales. Given the rapid change that the East Australian Current is experiencing, and its importance on driving seabirds' distributions, the methods applied in our study could be replicated in future analyses to assess possible spatial/temporal changes in seabird assemblages, including on its structure.
output: 
  # html_document: default
  # word_document:
  #   reference_docx: "../docs/docx-template.docx"
  # officedown::rdocx_document:
  #   reference_docx: "../docs/docx-template.docx"
  # rticles::sim_article:
  #   keep_tex: TRUE
  bookdown::word_document2:
    pandoc_args:
      - --lua-filter=../docs/scholarly-metadata.lua
      - --lua-filter=../docs/author-info-blocks.lua
      - --reference-doc=../docs/docx-template.docx 
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
---

@@

Seabird data were gathered aboard RV Investigator and RV Falkor during 15 voyages, between 2016--2021, over \~37° of latitude (**Fig. 1b**; **Table S1**). One observer continuously recorded seabirds within a 300-m forward quadrant (eye height **18-m, Eric?** a.s.l) throughout daylight hours while the ship was underway at a constant speed. Species, the number of birds and behaviour were logged directly onto a data entry system through a ship-based web server. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species) **(Eric: is there any ref for the protocol?)**.

@@

knitr::include_graphics("../results/Fig1_study-area-w-oceanogr.png")
```

@@

All models suggested two RCPs as the optimal grouping number, except for P/A-autumn, which indicated three RCPs as the best fit (**Fig S3. - multifit plots**). The best fitting models generally had **1--3** covariates selected, although the abundance-autumn model had five covariates (**Table 2 - full-model/selected model**). All the best models selected SST or SSS, variables essential for characterising water masses. The best models show a good fit with no apparent deviances in assumptions (**Fig. S4 - residuals**).

@@

@@

While remote tracking studies are advancing our understanding on how seabirds use their environment at fine-, individual-scale level [@tremblay2009], they fail to perceive how groups of species (assemblages) relate to the environment and each other on a shared seascape. Therefore, using at-sea counts of seabirds is a well-established method to capture these relationships [@ainley2012; @woehler2003; @hyrenbach2007]. On the other hand, the spatial scale on which assemblages relate to their environment is also a field of debate [@hunt1987; @ribic1997; @scales2017]. Understanding the biology of the studied group is thus essential to set the appropriate scale of analysis. Our choice of 1° latitude $\times$ 1° longitude was mainly influenced by data availability and the trade-off between number of species per grid/season (see also \@ref(limitations)); nonetheless, the chosen spatial scale is similar with seabird distribution analyses elsewhere [@harris2007; @santora2013; @sojitra2022]. Therefore, for describing seabird assemblages at a large scale, we believe the chosen spatial scale was a good fit, and models described expected patterns of species composition.
